\documentclass[b5paper]{report}

\input{preamble}

\begin{document}

Let \((X,\mathbb{P})\) be a probability space and \(F \colon X \times I \circlearrowleft\) a skew product defined by
\[F(x,y) = (g(x),f_x(y))\]
where \(g \colon X \circlearrowleft\) is a measure-preserving map.

For a bounded function \(\varphi \colon I \to \mathbb{R}\), we denote
\begin{equation*}
  U_{\varphi}(x,y) = \limsup_{n \to \infty}\frac{1}{n}\sum_{i = 0}^{n - 1}\varphi(\pi_2(F^i(x,y))).
\end{equation*}
Since \(\varphi\) is bounded, \(U_{\varphi}(x,y) = U_{\varphi}(F(x,y))\) everywhere. Indeed, since for any pair of sequences \((r_n)_n,(s_n)_n \subseteq \mathbb{R}\) we have
\begin{equation*}
  \lvert \limsup r_{n} - \limsup s_n \rvert \leq \limsup\, \lvert r_n - s_n \rvert,
\end{equation*}
we conclude that
\begin{align*}
  \lvert U_{\varphi}(x,y) - U_{\varphi}(F(x,y)) \rvert \le \limsup_{n \to \infty} \frac{1}{n} \lvert \varphi(y) - \varphi(\pi_2(F^n(x,y))) \rvert = 0.
\end{align*}

For a fixed bounded \(\varphi\), suppose the following is true:
\begin{hypothesis}\label{h:1}
  There exists a measurable function \(\ell \colon I \to \mathbb{R}\) and a family \(\{\Omega_y\}_{y \in I}\) of subsets of \(X\), each with full measure \(\mu(\Omega_y) = 1\), such that \(U_{\varphi}(x,y) = \ell(y)\) for all \(x \in \Omega_y\).
\end{hypothesis}

Even if \(U_{\varphi}\) is invariant for every point, it's not clear if \(\ell(f_x(y)) = \ell(y)\) a.e. (let alone for every \(y\)) because we're dealing with the set \(\{ (x,y) : g(x) \in \Omega_{f_{x}(y)} \}\) (which presumably could have small measure), so the argument of \cite{bonifantSchwarzianDerivativesCylinder2008} seems incomplete. If the pushforward of Lebesgue by the \(f_x\) are absolutely continuous to Lebesgue, however, we're good: 

\begin{lemma}\label{l1}
  Suppose \(f_xm \ll m\) for \(\mathbb{P}\)-a.e.\@ \(x \in X\). Then there exists a set \(\Omega \subseteq X \times I\) with full measure \((\mathbb{P}\otimes m)(\Omega) = 1\) where \(\ell(f_x(y)) = \ell(y)\) for all \((x,y) \in \Omega\).
\end{lemma}

\begin{proof}
  Let \[\Omega = \{ (x,y) : x \in \Omega_y \land g(x) \in \Omega_{f_x(y)} \}.\] It's clear from the definition of the \(\{ \Omega_y \}_{y \in I}\), combined with the fact \(U_{\varphi}\) is invariant everywhere, that \(\ell(f_x(y)) = \ell(y)\) in the set \(\Omega\). Also,
  \allowdisplaybreaks
  \begin{align*}
    (\mathbb{P} \otimes m)(\Omega^c) &\leq \underbrace{(\mathbb{P} \otimes m) (\{ (x,y) : x \notin \Omega_y \})}_{0\text{ by Fubini}} +\, (\mathbb{P} \otimes m)(\{ (x,y) : g(x) \notin \Omega_{f_x(y)} \})\\
                 &=  \int_X m (\{ y : g(x) \notin \Omega_{f_x(y)} \}) \;d \mathbb{P}(x)\\
                 &=  \int_X ((f_x)_*m) (\{ y : g(x) \notin \Omega_{y} \}) \;d \mathbb{P}(x).
  \end{align*}
  Since \(g_* \mathbb{P} = \mathbb{P}\), we have \[(\mathbb{P} \times m)(\{ (x,y) : g(x) \notin \Omega_y \}) =(\mathbb{P} \times m)(\{ (x,y) : x \notin \Omega_y \}) = 0.\]
  By Fubini, the function \(x \mapsto m(\{ y : g(x) \notin \Omega_y \})\) is \(\mathbb{P}\)-a.e.\@ equal to zero. Then the hypothesis \((f_x)_*m \ll m\) implies that \(x \mapsto (f_x)_*m(\{ y : g(x) \notin \Omega_y \})\) is zero \(\mathbb{P}\)-a.e., so the integral above is zero.
\end{proof}

\begin{corollary}
  If for \(\mathbb{P}\)-a.e. \(x \in X\) there is some \(\sigma\)-finite invariant measure \(\mu_x\) for \(f_x\) such that \(m \sim \mu_x\), then a set \(\Omega\) as in \cref{l1} exists.
\end{corollary}

\begin{proof}
  By Radon-Nikodym, write \(m = w \mu_x\). Then \((f_x)_*m = (w \circ f_x)\mu_x \ll m\).
\end{proof}

Now we will see the assumptions of the lemma are satisfied on the zero-Schwarzian context of \cite{bonifantSchwarzianDerivativesCylinder2008}, where each \(f_x \colon I \circlearrowleft\) is an orientation-preserving diffeomorphism with zero Schwarzian derivative. Indeed, this condition means that there exists a measurable function \(p \colon X \to \mathbb{R}\) such that each \(f_x = \gamma ^{-1} \circ h_x \circ \gamma\) is conjugate to the translation \(h_x(t) = t + p(x)\) via the change of coordinates
\begin{equation*}
  \gamma(y) = \log y - \log (1 - y).
\end{equation*}
Since the Lebesgue measure \(m_{\mathbb{R}}\) on \(\mathbb{R}\) is invariant for translations, the measure \(\mu = \gamma ^{-1}_* m_{\mathbb{R}}\) is invariant for every \(f_x\). It's also \(\sigma\)-finite, because \(\gamma\) is a measurable bijection, and \(m \sim \mu\) because \(\gamma ^{-1}\) is a diffeomorphism onto \((0,1)\) (thus locally Lipschitz).

The rest of the argument in \cite{bonifantSchwarzianDerivativesCylinder2008} is still true if \(\ell\) is invariant only in the full measure set \(\Omega\) (instead of all \((x,y)\) in a family of full-measure sets of \(x \in X\), indexed by \(y \in I\)). Note that the full conclusion of the Theorem 6.2 is also still true (for all \(y_0 \in (0,1)\)), because the function \(\ell\) is monotone.

\printbibliography

% Then we have the following:

% \begin{lemma}
%   Given \(\ell\), \(\{ \Omega_y \}_{y \in I}\) as in \cref{h:1}, there is another pair \(\tilde{\ell}, \{ \widetilde{\Omega}_y \}_{y \in I}\) that satisfies \cref{h:1} with the additional property that \(\tilde \ell(f_x(y)) = \tilde \ell(y)\) for all \(x \in \widetilde{\Omega}_y\).
% \end{lemma}

% \begin{proof}
%   Define \(\tilde \ell(x) = U_{\ell}(x,y)\).
% \end{proof}

\end{document}
